---
title: "Part B 503"
author: "Sarveshwari Singh"
date: "28/12/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


1. Summarising the number of hospital admissions for COPD in England between 2001 and 2010 - 

```{r}
# Reading the observed data 
England_o <- read_csv('/Users/sarveshwarisingh/Downloads/copdobserved.csv')
England_o
# Summarising observed data
summary(England_o)

# Reading the expected data
England_e <- read_csv('/Users/sarveshwarisingh/Downloads/copdexpected.csv')
# Summarising the expected data
England_e

# Reading the shapefiles
England_s <- readOGR(dsn='/Users/sarveshwarisingh/Downloads/ENGLAND SHAPEFILES', layer='englandlocalauthority')
England_s

# Creating a column with total observed cases in each local authority between 2001 and 2010
England_o$total_o <- rowSums(England_o[,2:11])
# Creating a column with total expected cases in each local authority between 2001 and 2010
England_e$total_e <- rowSums(England_e[,2:11])

# Creating a new dataset consisting of variables of interest - Name of authority, total observed and expected cases between 2001 and 2010
England_data <- data.frame(England_o$Name, England_o$total_o, England_e$total_e)

# Summarising this new dataset called 'England_data'
summary(England_data)
colnames(England_data)

# Cleaning for further analyses
England_data <- England_data %>%
  rename(Observed = England_o.total_o, Expected = England_e.total_e, Name = England_o.Name)
England_data <- England_data %>%
  rename(name= Name)
```

Summary - 

2. Estimating raw and smoothed standardised morbidity ratios (SMR). Spatial patterns and extreme values will be considered. 

```{r}
# Calculating raw SMRs
England_data$SMR_raw <- England_data$Observed/England_data$Expected
summary(England_data)

# Mapping the raw SMRs
SMR <- merge(England_s, England_data, by='name')
SMR <- st_as_sf(SMR)
ggplot(SMR, aes(fill=SMR_raw))+
  geom_sf(colour=NA)+
  theme_bw()+
  labs(x='Longitude', y='Latitude', fill='SMR')+
  scale_fill_gradientn(colours=brewer.pal(9, 'RdPu'), breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1.0, 1.2, 1.4, 1.6, 1.8, 2.0, 2.2))

# Calculating a smoothed SMR
# Creating a neighbourhood 
neigh <- poly2nb(England_s, row.names = rownames(England_s))
# Creating a matrix for following a function call
matr <- nb2mat(neigh, style='B')
#Running a smooth model
model <- S.CARleroux(formula = Observed ~ offset(log(Expected)), 
                     data=England_data, 
                     family = 'poisson', 
                     W = matr, 
                     burnin = 20000, 
                     n.sample = 100000, 
                     thin = 10, rho = 1)

England_data$SMR_smooth <- model$fitted.values/England_data$Expected
summary(England_data)

# Mapping the smooth SMRs

SMR <- merge(England_s, England_data, by='name')
SMR <- st_as_sf(SMR)
ggplot(SMR, aes(fill=SMR_smooth))+
  geom_sf(colour=NA)+
  theme_bw()+
  labs(x='Longitude', y='Latitude', fill='SMR')+
  scale_fill_gradientn(colours=brewer.pal(9, 'RdPu'), breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1.0, 1.2, 1.4, 1.6, 1.8, 2.0, 2.2))

# Plotting both raw and smooth SMRs together

SMR_both <- reshape2::melt(England_data[,c("name", "SMR_raw", "SMR_smooth")],
                           id.vars = "name",
                           variable.name = 'Type',
                           value.name = 'SMR')

SMR_both$Type <- as.character(SMR_both$Type)
SMR_both$Type[SMR_both$Type == 'SMR_raw'] <- 'Raw SMRs'
SMR_both$Type[SMR_both$Type == 'SMR_smooth'] <- 'Smoothed SMRs'

# Combining Raw SMRs and the shapefile
SMR_both <- merge(st_as_sf(England_s),
                  SMR_both,
                  by = 'name')

# Creating map of Raw and smoothed SMRs
ggplot(SMR_both,
       aes(fill = SMR)) +
  # Plot as a map using the shapefiles
  geom_sf(colour = NA) +
  # Black and white theme
  theme_bw() +
  # Plot labels
  labs(x = 'Longitude',
       y = 'Latitude',
       fill = 'SMR') +
  # Facetting by SMR type
  facet_wrap(. ~ Type) +
  # Setting bands and colour scheme on the legend
  scale_fill_gradientn(colours = brewer.pal(9, 'RdPu'),
                       breaks = c(0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1.0, 1.1, 1.2, 1.3, 1.4, 1.5, 1.6, 1.7, 1.8, 1.9, 2.0, 2.1, 2.2))


```

Spatial patterns - In England, hospital admission rates appear to be higher than the national average for the northern, north-west, north-east, Yorkshires, and the south-west regions. From the spatial mapping it is clear that the overall distribution of risk is stable. There seem to be no significant differences between raw and smoothed SMRs in local authorities, however this will be tested in the following analysis. 

Extreme values 

```{r}
# some exploratory numerical analysis
England_data[England_data$SMR_smooth == min(England_data$SMR_smooth), ]  
England_data[England_data$SMR_raw == min(England_data$SMR_raw), ] 
England_data[England_data$SMR_raw == max(England_data$SMR_raw), ] 
England_data[England_data$SMR_smooth == max(England_data$SMR_smooth), ] 
ab_difference <- sort(abs(England_data$SMR_smooth - England_data$SMR_raw))
ab_difference[length(ab_difference)]
plot(ab_difference)
England_data[abs(England_data$SMR_smooth - England_data$SMR_raw) == ab_difference[length(ab_difference)],]
England_data[abs(England_data$SMR_smooth - England_data$SMR_raw) == ab_difference[length(ab_difference)-1],]
England_data[abs(England_data$SMR_smooth - England_data$SMR_raw) == ab_difference[length(ab_difference)-2],]

```

Extreme values - 
Are the big changes appropriate? - 

3. Changes in the risks of hospitalisation over time in England are explored. 

```{r}
# How SMR changes over time
max(England_e[,11])
Observed_ts <- colSums(England_o[,c(2:11)])
Observed_ts
Expected_ts <- colSums(England_e[, c(2:11)])
Expected_ts
SMR_ts <- Observed_ts/Expected_ts
SMR_ts_df <- as.data.frame(SMR_ts)
row.names(SMR_ts_df) <- c("2001", "2002", "2003", "2004", "2005", "2006", "2007", "2008", "2009", "2010")
SMR_ts_df$Date <- c("2001", "2002", "2003", "2004", "2005", "2006", "2007", "2008", "2009", "2010")
SMR_ts_df
ggplot(data=SMR_ts_df, aes(x=Date, y=SMR_ts, group=1))+
  geom_line()
```

